A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer

### 2024-01-18 content/manuscript.qmd
add info about data & edits (see below)
I have removed the formalism related to sensitivity and specificity.
Also, the mixture model is now not motivated as a lift of the old
model's limitations anymore.

---
title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliations:
  - id: 1
    name: University of Zurich
    departement: Departement of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    departement: Departement of Radiation Oncology
    city: Zurich
    country: Switzerland
bibliography: ../references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
format:
  iccr-pdf:
    keep-tex: true
    include-in-header:
      - text: "\\usepackage{newtxtext,newtxmath}"
  html:
    code-line-numbers: true
    code-tools: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

HNSCC patients with primary tumors at different subsites, e.g. in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999]. To naively capture this in a risk prediction using our previously developed model, one would need to train entirely separate models for each subsite, based on separated datasets. We consider this a limitation for two reasons:

1. While the spread from the tumor to the LNLs may differ for different primary tumor subsites, we would assume the spread from involved LNLs further down the lymphatic network to still be similar. Thus, a dataset of oral cavity patients may contain information that is relevant to the oropharynx model and vice versa.
2. A tumor may lie anatomically between two subsites. Then, it makes sense to assume that this tumor spreads in a way that can be described as a combination of the typical spread patterns of these two subsites, given they are close enough to each other. This presents an opportunity to reduce the model's complexity: If a subsite's lymphatic spread can be described as a mixture of two others' spread patterns, we only require two distinct HMMs to describe three tumor subsites.

In the work presented here, we implement a mixture of HMMs that aims to address these limitations. In @sec-materials-and-methods we first introduce the basic concept of the HMM (@sec-hmm), followed by a general description of mixture models (@sec-mixture-of-hmms) and how we implemented them in our particular case.

@@

![Parametrized graph representation of the lymphatic network considering four LNLs. Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs. Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="22%"}

@@

Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a hidden binary random variable (RV) $X_v[t]$ taking on the true -- but not directly observable -- state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved"). Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's hidden state w.r.t. their lymphatic involvement at time $t$.

When a patient gets diagnosed, their LNLs' involvement status is assessed using a clinical modality $\mathcal{O}$ with its repsective sensitivity $s_N^\mathcal{O}$ and specificity $s_P^\mathcal{O}$. This involvement status, too, is modelled as an RV $Y_v^\mathcal{O}$. Hence, a diagnosis obtained with diagnostic modality $\mathcal{O}$ can be collected in the random vector $\mathbf{Y}^\mathcal{O}$.

We model the process of tumor progression via lymphatic drainage by connecting the hidden RVs in a graph, as shown in @fig-graph-with-obs. The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities. The orange arcs, as mentioned, correspond to observing a diagnosis $Z_v^\mathcal{O}$, given a hidden state $X_v$. The values this conditional probability $P(Y_v \mid X_v)$ can take on are tabulated in @tbl-sens-spec.

+---------+---------+---------+
|         | $X_v=0$ | $X_v=1$ |
+=========+:=======:+:=======:+
| $Y_v=0$ | $s_P$   | $1-s_N$ |
+---------+---------+---------+
| $Y_v=1$ | $1-s_P$ | $s_N$   |
+---------+---------+---------+

: Table encoding the conditional probability $P(Y_v \mid X_v)$ to correctly (or wrongly) diagnose an LNL's true hidden state. This table is associated with the orange arcs in @fig-graph-with-obs. {#tbl-sens-spec}

Similarily, the red arcs in @fig-graph-with-obs depict the probability that the primary tumor spreads within one abstract time-step. Lastly, the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level. For example, the edge from $X_2$ to $X_3$ encodes the probability $P\left( X_3[t+1] \mid X_2[t] \right)$, which is parametrized with $b_3$, and $t_{2 \rightarrow 3}$ and tabulated in @tbl-spread-2-to-3. There is an additional restriction on any LNL's state $X_v[t+1]=0$ to be healthy: It requires that the level was also healthy in the time-step before, meaning $X_v[t]=0$. This is because we assume no spontaneous self-healing of metastatic levels.

@@

With the introduced conditional probabilities, we can now compute the joint probability of any complete hidden state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitioning to any other possible state $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$. Here, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs. In the graph shown in @fig-graph-with-obs, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$. Because these terms are essentially products of terms like those in @tbl-spread-2-to-3. We can then collect these terms in a _transition matrix_ $\mathbf{A}$:

@@

To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the sums in the denominator of @eq-bayes-law:

$$
P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
$$ {#eq-likelihood}

@@

@@

Introducing this latent variable is helpful, because the mixture model likelihood is invariant w.r.t. permutations of the component labels. This introduces a number of problems, e.g. for common MCMC sampling methods. The latent variable $\mathbf{Z}$ allows us to derive two sets of interdependent equations that we may solve in an iterative fashion (see e.g. @bishop_pattern_2006 for a detailed derivation) that is commonly referred to as _expectation-maximization (EM)_ algorithm:

The first are the probabilities of subcohort $i$ to belong to component $j$, given a set of parameters $\boldsymbol{\Psi}^\star$. These are often called the _responsibilities_:
$$
\gamma (z_{ij}) = P \left( z_{ij}=1 \mid \mathbf{D}_i , \boldsymbol{\Psi}^\star, \mathbf{c} \right) = \frac{ c_j P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_j^\star \right) }{\sum_{k=0}^g c_k P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_k^\star \right)}
$$

From this, we can compute new mixing proportions $c_j^\star=\sum_{i=1}^s \gamma (z_{ij}) / s$ and then infer new parameters $\boldsymbol{\Psi}^\star$ -- e.g. via MCMC sampling -- from the resulting likelihood:

@@

>>>>

---
title: A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer
author:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julian Brönnimann
    affiliations:
      - ref: 1
      - ref: 2
  - name: Yoel Perez Haas
    affiliations:
      - ref: 1
      - ref: 2
  - name: Esmée Lauren Looman
    affiliations:
      - ref: 1
      - ref: 2
  - name: Sergi Benavente
    orcid: 0000-0002-2700-3522
    affiliations:
      - ref: 11
  - name: Adrian Schubert
    affiliations:
      - ref: 3
      - ref: 4
      - ref: 7
  - name: Dorothea Barbatei
    affiliations:
      - ref: 8
  - name: Laurence Bauwens
    affiliations:
      - ref: 8
  - name: Jean-Marc Hoffmann
    orcid: 0000-0003-1193-0863
    affiliations:
      - ref: 2
  - name: Sandrine Werlen
    affiliations:
      - ref: 4
      - ref: 5
  - name: Olgun Elicin
    orcid: 0000-0002-6996-0646
    affiliations:
      - ref: 3
  - name: Matthias Dettmer
    orcid: 0000-0003-0948-1392
    affiliations:
      - ref: 6
      - ref: 10
  - name: Philippe Zrounba
    orcid: 0000-0001-6223-6113
    affiliations:
      - ref: 9
  - name: Bertrand Poumayou
    orcid: 0000-0002-8060-7827
    affiliations:
      - ref: 2
  - name: Panagiotis Balermpas
    orcid: 0000-0001-5261-6446
    affiliations:
      - ref: 2
  - name: Vincent Grégoire
    affiliations:
      - ref: 8
  - name: Roland Giger
    orcid: 0000-0002-5574-3210
    affiliations:
      - ref: 4
      - ref: 5
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    affiliations:
      - ref: 1
      - ref: 2
affiliations:
  - id: 1
    name: University of Zurich
    department: Department of Physics
    city: Zurich
    country: Switzerland
  - id: 2
    name: University Hospital Zurich
    department: Department of Radiation Oncology
    city: Zurich
    country: Switzerland
  - id: 3
    name: Bern University Hospital
    department: Department of Radiation Oncology
    city: Bern
    country: Switzerland
  - id: 4
    name: Bern University Hospital
    department: Department of ENT, Head & Neck Surgery
    city: Bern
    country: Switzerland
  - id: 5
    name: Bern University Hospital
    department: Head and Neck Anticancer Center
    city: Bern
    country: Switzerland
  - id: 6
    name: Bern University Hospital
    department: Institute of Tissue Medicine and Pathology
    city: Bern
    country: Switzerland
  - id: 7
    name: Réseau Hospitalier Neuchâtelois
    department: Department of ENT, Head & Neck Surgery
    city: Neuchâtelois
    country: Switzerland
  - id: 8
    name: Centre Léon Bérard
    department: Department of Radiation Oncology
    city: Lyon
    country: France
  - id: 9
    name: Centre Léon Bérard
    department: Department of Head and Neck Surgery
    city: Lyon
    country: France
  - id: 10
    name: Klinikum Stuttgart
    department: Institute of Pathology
    city: Stuttgart
    country: Germany
  - id: 11
    name: Hospital Vall d'Hebron
    department: Departement of Radiation Oncology
    city: Barcelona
    country: Spain
bibliography: ../references.bib
abstract: |
  We previously developed a mechanistic hidden Markov model (HMM) to predict the
  lymphatic tumor progression in oropharyngeal squamous cell carcinomas (OPSCCs). To
  extend the model to other tumor subsites, defined by ICD-10 codes, in the head and
  neck, we employ a mixture of these HMMs and learn the cluster assignments and model
  parameters in an iterative, EM-like algorithm from multi-centric data. The mixture
  model manages to cluster anatomically close subsites and correctly infers the
  clusters' parameters. Using this mixture model allows the prediction of individual
  risks of occult nodal disease, given a diagnosis that includes tumor subsite.
format:
  iccr-pdf:
    keep-tex: true
    include-in-header:
      - text: "\\usepackage{newtxtext,newtxmath}"
  html:
    code-line-numbers: true
    code-tools: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

HNSCC patients with primary tumors at different subsites, e.g. in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999]. Our model does so far not have the capability to naturally describe different tumor subsites. To that end, we present an approach using mixtures of these HMMs in this work. This makes intuitive sense, because if a tumor lies anatomically between two anatomically close subsites with slightly different spread patterns, we may be able to describe its lymphatic progression as a mixture of these two spread patterns.

@@

@@

Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable (RV) $X_v[t]$ taking on the true state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved"). Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's state w.r.t. their lymphatic involvement at time $t$.

We model the process of tumor progression via lymphatic drainage by connecting the RVs in a graph, as shown in @fig-graph-with-obs. The arcs in [this graph](#fig-graph-with-obs) represent conditional probabilities. The orange arcs correspond to observing a diagnosis $Y_v$, given the true state $X_v$. For the sake of brevity, we will not go into the details of describing how to infer the true -- but technically hidden -- state of LNL involvement from diagnoses with lower-than-one sensitivity and specificity. This description can be found in @zora231470. Throughout this work, we instead combine diagnostic and pathologic involvement information from the data into a "maximum likelihood" diagnosis and assume its sensitivity and specificity to be one, meaning the normally hidden state $X_v$ becomes the observed state. This simplification is reasonable for pathologic involvement and because we are at this stage more interested in testing whether our model is able to describe a realistic probability distribution over lymphatic involvement.

The red arcs in the graph of @fig-graph-with-obs depict the probability that the primary tumor spreads within one abstract time-step. While the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level. For example, the edge from $X_2$ to $X_3$ encodes the probability $P\left( X_3[t+1] \mid X_2[t] \right)$, which is parametrized with $b_3$, and $t_{2 \rightarrow 3}$ and tabulated in @tbl-spread-2-to-3. There is an additional restriction on any LNL's state $X_v[t+1]=0$ to be healthy: It requires that the level was also healthy in the time-step before, meaning $X_v[t]=0$. This is because we assume no spontaneous self-healing of metastatic levels.

@@

With the introduced conditional probabilities, we can now compute the joint probability of any complete state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitioning to any other possible state $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$. Here, when we use $\boldsymbol{\xi}$ instead of $\mathbf{x}$ for the values the random vector $\mathbf{X}$ can take on, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs. In the graph shown in @fig-graph-with-obs, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$. Because these terms are essentially products of terms like those in @tbl-spread-2-to-3. We can then collect these terms in a _transition matrix_ $\mathbf{A}$:

@@

To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{x}_1, \mathbf{x}_2, \ldots, \mathbf{x}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the terms in @eq-marginalized-evolution:

$$
P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N P \left( \mathbf{X} = \mathbf{x}_i \mid T_i, \boldsymbol{\theta} \right)
$$ {#eq-likelihood}

@@

![Parametrized graph representation of the lymphatic network considering four LNLs. Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs. Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="25%" fig-pos="th"}

@@

The latent variables are helpful in resolving the invariance of the likelihood w.r.t. permutations of the component labels, which may introduce problems, e.g. for common MCMC sampling methods. The $\mathbf{Z}$ allows us to derive two sets of interdependent equations that we may solve in an iterative fashion (see e.g. @bishop_pattern_2006 for a detailed derivation) that is commonly referred to as _expectation-maximization (EM)_ algorithm:

The first set are the probabilities of subcohort $i$ to belong to component $j$, given a set of parameters $\boldsymbol{\Psi}^\star$. These are often called the _responsibilities_:
$$
\gamma (z_{ij}) = P \left( z_{ij}=1 \mid \mathbf{D}_i , \boldsymbol{\Psi}^\star, \mathbf{c} \right) = \frac{ c_j P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_j^\star \right) }{\sum_{k=0}^g c_k P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_k^\star \right)}
$$

From this, we can compute new mixing proportions $c_j^\star=\sum_{i=1}^s \gamma (z_{ij}) / s$ and then infer new parameters $\boldsymbol{\Psi}^\star$ -- e.g. via MCMC sampling -- from the resulting likelihood, which is the second set:

@@

## Multi-Centric Data

For the analyses in this work, we used five datasets from four different institutions:

1. 287 oropharyngeal patients from the University of Zurich in Swizerland
2. 263 oropharyngeal patients from the Centre Léon Bérard in Fance
3. 289 oropharyngeal and oral cavity patients from the Inselspital Bern in Swizerland
4. 239 oropharyngeal and oral cavity patients from the Centre Léon Bérard in Fance
5. 162 oropharyngeal patients from the Hospital Vall d'Hebron in Spain

The data comes in the form of CSV tables and are -- except for the last and most recent addition -- publicly available [@ludwig_detailed_2022;@ludwig_multi-centric_2023] and may be interactively explored in our **Ly**mphatic **Pro**gression e**X**plorer [LyProX](https://lyprox.org). Each row of these tables corresponds to one patient and details in which LNL metastatic involvement was found or not, according to different diagnostic and pathologic modalities.
